The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the Last Glacial Maximum

### 2013-08-23 sections/climate.tex
Improve climate part and add missing references

We drive the mass balance routine of PISM with temperature and precipitation data from one observational dataset and climatologies from three global and one regional reanalyses. We here present the different climate datasets used in this study.

@@

WorldClim is a high-resolution climate dataset built from meteorological observations \citep{data:worldclim}. These climate surfaces were generated by spatial interpolation, using 30\,arc\,s-aggregated hole-filled SRTM elevation data \citep{jarvis-etal-2008}, between a large set of measurements taken at weather stations over global land areas, using . This provides a resolution much higher than attained through GCM in areas where sufficient amount of weather stations exist.

 In the study area, the spatial distributions of summer (JJA) and winter (DJF) air surface temperatures in WorldClim generally show a decrease from south to north (JJA) and from southwest to northeast (DJF), such that regions further inland experience colder winters (Fig.~\ref{fig:temp}). It should be noted that temperature gradients are reversed (colder inland than at the coast) and much stronger in winter than in summer. Temperatures rises well above zero during the summer months over the entire modelling domain, except for the highest mountain peaks, and even regions where mean annual temperatures are well below freezing do experience warm summers. In other words, there is a strong contrast in seasonality between coastal temperate-climate and inland continental-climate regions.

@@

There are two problems inherent to the use of WorldClim data in our region of study. Firstly, the density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous. Although good coverage exists for the southern part of our modelling domain, several hundred kilometres can separate neighbouring stations in the north \citep{data:worldclim}. Secondly, WorldClim lacks of data offshore, which would have been particularly useful for the on the Pacific continental shelf which was partly covered by ice during the LGM \citep{jackson-clague-1991}.

@@

In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global climate reanalyses and one regional climate reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate System Forecast Reanalysis (CFSR), and the North American Regional Reanalysis (NARR). Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the NOAA/OAR/ESRL Physical Sciences Division, Boulder, Colorado, USA, from their website at \url{http://www.esrl.noaa.gov/psd/} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table \ref{tab:reanalyses}).

The spatial distributions of JJA air surface temperatures and DJF precipitation rates from the four reanalyses datasets are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}. Summer temperature and winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation.

Because reanalysis are a product combinaing observations and circulation modelling, we believe that the input from reanalyses may perform better than WorldClim in sparsely monitored regions such as the northernmost American Cordillera.

@@

WorldClim, NCEP/NCAR, ERA-Interim, CFSR and NARR climatologies were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to a 10\,km model grid (see section~\ref{sec:model}). In addition, WorldClim data was extrapolated to cover oceans grid points using a nearest-neighbour approach. These last two steps are not presented on figures~\ref{fig:temp} and~\ref{fig:prec}.

An alternative forcing was prepared for CFSR to correct for artefacts seen in its precipitation field (Fig.~\ref{fig:prec}). A smoothing of the precipitation field was achieved by averaging data locally in a circular kernel of 7\,pixels in diameter prior to re-projection. These preprocessing steps were performed in GRASS~GIS.

Throughout the simulation, the numerical model dynamically applies temperature lapse-rate corrections that account for differences between the climate forcing reference topography and the ice flow model topography on the one hand, and the evolution of ice thickness on the over hand. For reanalysis data, the GCM surface topography (or surface geopotential height) was used as reference topography. For the WorldClim data, the aggregated hole-filled SRTM altitude field is used as reference topography. As a basal condition for the ice flow model, we use the ETOPO1\citep{data:etopo1} combined topography and bathymetry dataset. The reference topographies from the five forcing datasets, alongside the ETOPO1 data are shown in figure~\ref{fig:topo}. An annual lapse-rate of 6\unit{\degree C\,km^{-1}} is used in all simulations. No lapse-rate corrections apply to precipitation rates.

>>>>

Our atmospheric forcing consists of monthly climatologies of surface air temperature and precipitation obtained from one observational dataset, three global reanalysis and one regional reanalysis.

@@

WorldClim is a high-resolution climate dataset built from meteorological observations \citep{data:worldclim}. It was built by spatial interpolation over global land areas, using 30\,arc\,s-aggregated hole-filled SRTM elevation data \citep{data:srtm} and GTOPO30 elevation data\citep{data:gtopo30}, between a large set of measurements taken at weather stations worldwide. This provides a resolution much higher than attained through GCM in areas where sufficient amount of observations exist.

Within our modelling domain, the spatial distributions of summer (JJA) and winter (DJF) air surface temperatures in WorldClim generally show a decrease from south to north (JJA) and from south-west to north-east (DJF), such that regions further inland experience colder winters (Fig.~\ref{fig:temp}). It should be noted that temperature gradients are reversed (colder inland than at the coast) and much stronger in winter than in summer. Temperatures rises well above zero during the summer months over the entire modelling domain, except for the highest mountain peaks, and even regions where mean annual temperatures are well below freezing do experience warm summers. In other words, there is a strong contrast in seasonality between coastal temperate-climate and inland continental-climate regions.

@@

There are two problems inherent to the use of WorldClim data in our study area. Firstly, the density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous. Although good coverage exists for the southern part of our modelling domain, several hundred kilometres can separate neighbouring stations in the north \citep{data:worldclim}. Secondly, WorldClim lacks of data offshore, which would have been particularly useful over the Pacific continental shelf which was partly covered by ice during the LGM \citep{jackson-clague-1991}.

@@

In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global climate reanalyses and one regional climate reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis\citep{data:ncar}, the ERA-Interim reanalysis\citep{data:erai}, the Climate System Forecast Reanalysis (CFSR, \citet{data:cfsr}), and the North American Regional Reanalysis (NARR, \citet{data:narr}). Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the \citet{web:psd} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table \ref{tab:reanalyses}).

The spatial distributions of JJA air surface temperatures and DJF precipitation rates from the four reanalyses climatologies are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}. Summer temperature and winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation.

Because reanalyses include observational information through data assimilation, they are more accurate than simple GCMs in densely monitored regions, while offering physically based output in sparsely monitored regions such as the northern part of our modelling domain (Bengtsson et al., 2007).

@@

WorldClim, NCEP/NCAR, ERA-Interim, CFSR and NARR climatologies were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the 10\,km-resolution model grid. In addition, WorldClim data was extrapolated to cover oceans grid points using a nearest-neighbour approach. Note that we choose to present here (Fig.~\ref{fig:temp} and~\ref{fig:prec}) the original rather than interpolated data to emphasize differences in spatial resolution among datasets.

An alternative forcing was prepared for CFSR to correct for artefacts seen in its precipitation field (Fig.~\ref{fig:prec}). A smoothing of the precipitation field was achieved by averaging data locally in a circular kernel of 7\,pixels in diameter prior to re-projection. These preprocessing steps were performed in GRASS~GIS using Python packages Numpy, Scipy and NetCDF4-Python\needref.
